---
title: "hg38 ↔ T2T-CHM13 Transposable Element Liftover Analysis"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    theme: flatly
    toc: true
    toc_float: true
    toc_depth: 3
    number_sections: true
    df_print: paged
  pdf_document:
    toc: true
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
library(knitr)
library(kableExtra)
library(dplyr)

data_dir <- file.path(dirname(knitr::current_input(dir = TRUE)), "results")

# Pre-load totals used in captions and the data guide
.t1   <- read.delim(file.path(data_dir, "t1_class_counts.tsv"))
.hg38_total   <- .t1$hg38_n[.t1$class  == "Total"]
.chm13_total  <- .t1$chm13_n[.t1$class == "Total"]
.hg38_sine_n  <- .t1$hg38_n[.t1$class  == "SINE"]
.chm13_sine_n <- .t1$chm13_n[.t1$class == "SINE"]
.hg38_line_n  <- .t1$hg38_n[.t1$class  == "LINE"]
.chm13_line_n <- .t1$chm13_n[.t1$class == "LINE"]

.t6              <- read.delim(file.path(data_dir, "t6_liftover_both.tsv"))
.hg38_unmap_all  <- .t6$unmapped[1]
.chm13_unmap_all <- .t6$unmapped[2]

.t9              <- read.delim(file.path(data_dir, "t9_linesine_comparison.tsv"))
.hg38_ls_total   <- .t9$source_linesine_total[1]
.chm13_ls_total  <- .t9$source_linesine_total[2]

.t5    <- read.delim(file.path(data_dir, "t5_alu_age.tsv"))
.tot5  <- .t5[.t5$age_group == "Total", ]

N <- function(n) format(n, big.mark = ",", scientific = FALSE)

kt <- function(df, caption = NULL, ...) {
  kable(df, caption = caption, format.args = list(big.mark = ","),
        align = c("l", rep("r", ncol(df) - 1)), ...) %>%
    kable_styling(bootstrap_options = c("striped", "hover", "condensed", "bordered"),
                  full_width = FALSE, position = "left") %>%
    row_spec(nrow(df), bold = TRUE, background = "#f0f0f0")
}
```

> **Motivation:** Are there more assembly-specific TE loci in hg38 than in T2T-CHM13?
> And are young, recently-mobile Alu elements disproportionately absent from
> the other assembly — evidence that they lack stable genomic addresses across individuals?

---

# Data Sources & Table Guide {#data-sources}

## Input files

Two RepeatMasker annotation files were used throughout this analysis.
Every table in this report is derived from one or both of them.

| File | Assembly | Format | Size | Origin |
|------|----------|--------|------|--------|
| `hg38_rmsk.txt.gz` | hg38 (GRCh38) | UCSC rmsk.txt | 149 MB compressed | UCSC Genome Browser, Oct 2022 |
| `T2T-CHM13v2.0_RepeatMasker.bed` | T2T-CHM13v2.0 (hs1) | BED | 327 MB uncompressed | project `references/annotations/` |

The hg38 file was downloaded specifically for this analysis and is stored in
`analysis/liftover/data/`. The CHM13 file was already in the project repository
(used by the SoloTE pipeline) and was **not** re-downloaded.

## Table guide: what each table shows and how percentages are calculated

```{r guide-table}
guide <- data.frame(
  Table = c(
    "Table 1",
    "Table 2",
    "Table 3",
    "Table 6",
    "Table 4a",
    "Table 4b",
    "Table 7",
    "Table 8",
    "Table 9",
    "Table 5"
  ),
  `Source file(s)` = c(
    "both files",
    "both files",
    "both files",
    "hg38 file (row 1); CHM13 file (row 2)",
    "hg38 file",
    "hg38 file",
    "CHM13 file",
    "CHM13 file",
    "both files",
    "hg38 file (background); liftover outputs (unmapped sets)"
  ),
  `Denominator (n=)` = c(
    paste0("hg38: ", N(.hg38_total), "; CHM13: ", N(.chm13_total)),
    paste0("hg38 SINEs: ", N(.hg38_sine_n), "; CHM13 SINEs: ", N(.chm13_sine_n)),
    paste0("hg38 LINEs: ", N(.hg38_line_n), "; CHM13 LINEs: ", N(.chm13_line_n)),
    paste0("hg38: ", N(.hg38_total), "; CHM13: ", N(.chm13_total)),
    paste0(N(.hg38_unmap_all), " hg38-specific unmapped loci"),
    paste0("top 20 of ", N(.hg38_unmap_all), " hg38-specific unmapped"),
    paste0(N(.chm13_unmap_all), " CHM13-specific unmapped loci"),
    paste0("top 20 of ", N(.chm13_unmap_all), " CHM13-specific unmapped"),
    paste0("hg38 LINE/SINE: ", N(.hg38_ls_total), "; CHM13 LINE/SINE: ", N(.chm13_ls_total)),
    paste0("hg38 Alu bg: ", N(.tot5$hg38_bg_n),
           "; hg38-specific Alus: ", N(.tot5$hg38_unmap_n),
           "; CHM13-specific Alus: ", N(.tot5$chm13_unmap_n))
  ),
  `% column means` = c(
    "fraction of all annotations in that assembly",
    "fraction of all SINEs in that assembly",
    "fraction of all LINEs in that assembly",
    "fraction of source loci that failed liftover",
    "col 3: fraction of 58,933 unmapped; col 4: fraction of all hg38 loci in that class",
    "fraction of all 58,933 hg38-specific unmapped loci",
    "col 3: fraction of 89,234 unmapped; col 4: fraction of all CHM13 loci in that class",
    "fraction of all 89,234 CHM13-specific unmapped loci",
    "fraction of source LINE/SINE loci that failed liftover",
    "each % column is independent: fraction within that column's total"
  ),
  check.names = FALSE
)

kable(guide, caption = "Table guide — data sources and percentage denominators",
      align = c("l", "l", "l", "l")) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "bordered"),
                full_width = TRUE, position = "left") %>%
  column_spec(1, bold = TRUE, width = "6em") %>%
  column_spec(3, width = "22em") %>%
  column_spec(4, width = "26em")
```

---

# How CrossMap Liftover Works {#how-crossmap}

## What is a liftover?

A **liftover** translates genomic coordinates from one assembly to another.
Given a locus like `chr1:1,000,000–1,000,300` in hg38, liftover asks: *what is
the equivalent position in T2T-CHM13?*

CrossMap v0.7.0 implements this using **chain files** — precomputed pairwise
alignments from the UCSC Genome Browser. A chain file encodes which hg38 blocks
are syntenic to (i.e., evolutionarily conserved and unambiguously aligned to)
which CHM13 blocks. Two chain files were used here:

| Chain file | Direction |
|------------|-----------|
| `hg38ToHs1.over.chain.gz` | hg38 → CHM13 (hs1) |
| `hs1ToHg38.over.chain.gz` | CHM13 → hg38 |

## What does "hg38 → CHM13" mean?

The direction **hg38 → CHM13** means: take each locus annotated in the hg38 file
(`hg38_rmsk.txt.gz`) and ask whether it has a syntenic counterpart in CHM13
coordinate space.

- **Mapped** → the locus has a defined syntenic position in CHM13. It exists in
  both assemblies (though possibly at a slightly different coordinate).
- **Unmapped** → the locus has **no unambiguous syntenic counterpart** in CHM13.
  Unmapped covers three distinct situations:
  1. *True absence* — the sequence genuinely does not exist in the CHM13 individual's
     genome (a polymorphic insertion absent from CHM13).
  2. *Assembly difference* — the region is collapsed, duplicated, or resolved
     differently between the two assemblies (e.g., centromere expansion in T2T).
  3. *Ambiguous multi-hit* — the locus maps to **multiple** positions in CHM13.
     Unlike STAR (which reports all mapping positions), CrossMap cannot assign the
     locus to a single coordinate and therefore calls it **unmapped**. This is
     analogous to a read that multi-maps in STAR and is excluded from unique-mapper
     counts.

## Why run both directions?

Running hg38 → CHM13 and CHM13 → hg38 independently gives two distinct
assembly-specific sets. Each uses a different source file:

| Direction | Source file | Unmapped set |
|-----------|-------------|--------------|
| hg38 → CHM13 | `hg38_rmsk.txt.gz` | hg38-specific loci — present in hg38 but absent/unresolved in CHM13 |
| CHM13 → hg38 | `T2T-CHM13v2.0_RepeatMasker.bed` | CHM13-specific loci — present in CHM13 but absent/unresolved in hg38 |

These sets are **not mirror images**. They differ because the two assemblies have
different overall structures: T2T-CHM13 resolves centromeres and other difficult
regions that are collapsed in hg38, while hg38 may contain sequences absent from
the single CHM13 individual.

---

# Total RepeatMasker Annotations {#total}

**Source:** both files. Percentages are the fraction of all annotations
in that assembly (`hg38 %` = count / `r N(.hg38_total)` total hg38 loci;
`CHM13 %` = count / `r N(.chm13_total)` total CHM13 loci).

```{r table1}
t1 <- read.delim(file.path(data_dir, "t1_class_counts.tsv"))
t1 <- t1 %>%
  mutate(
    hg38_pct  = ifelse(class == "Total", "100%",
                  paste0(round(hg38_n  / .hg38_total  * 100, 1), "%")),
    chm13_pct = ifelse(class == "Total", "100%",
                  paste0(round(chm13_n / .chm13_total * 100, 1), "%"))
  ) %>%
  select(Class = class,
         `hg38 count` = hg38_n,
         `hg38 %` = hg38_pct,
         `CHM13 count` = chm13_n,
         `CHM13 %` = chm13_pct)

kt(t1, caption = paste0(
  "Table 1. Total TE annotations by class  ",
  "(hg38 n = ", N(.hg38_total), ";  CHM13 n = ", N(.chm13_total), ")"
))
```

---

# SINE and LINE Family Breakdown {#families}

**Source:** both files. The hg38 counts come from `hg38_rmsk.txt.gz`
(`r N(.hg38_total)` total annotations) and the CHM13 counts come from
`T2T-CHM13v2.0_RepeatMasker.bed` (`r N(.chm13_total)` total annotations).
These two files were obtained independently and are not derived from one another.
Percentages within each table are the fraction of that assembly's loci
in the corresponding class (i.e. hg38 SINE % = count / `r N(.hg38_sine_n)`
total hg38 SINEs).

## SINE Families

```{r table2}
t2 <- read.delim(file.path(data_dir, "t2_sine_families.tsv"))
total_row2 <- t2[t2$family == "Total", ]
t2 <- t2 %>%
  mutate(
    hg38_pct  = ifelse(family == "Total", "100%",
                  paste0(round(hg38_n  / total_row2$hg38_n  * 100, 1), "%")),
    chm13_pct = ifelse(family == "Total", "100%",
                  paste0(round(chm13_n / total_row2$chm13_n * 100, 1), "%"))
  ) %>%
  select(Family = family,
         `hg38 count` = hg38_n,
         `hg38 % of SINEs` = hg38_pct,
         `CHM13 count` = chm13_n,
         `CHM13 % of SINEs` = chm13_pct)

kt(t2, caption = paste0(
  "Table 2. SINE family breakdown  ",
  "(hg38 total SINEs n = ", N(.hg38_sine_n),
  ";  CHM13 total SINEs n = ", N(.chm13_sine_n), ")"
))
```

## LINE Families

```{r table3}
t3 <- read.delim(file.path(data_dir, "t3_line_families.tsv"))
total_row3 <- t3[t3$family == "Total", ]
t3 <- t3 %>%
  mutate(
    hg38_pct  = ifelse(family == "Total", "100%",
                  paste0(round(hg38_n  / total_row3$hg38_n  * 100, 1), "%")),
    chm13_pct = ifelse(family == "Total", "100%",
                  paste0(round(chm13_n / total_row3$chm13_n * 100, 1), "%"))
  ) %>%
  select(Family = family,
         `hg38 count` = hg38_n,
         `hg38 % of LINEs` = hg38_pct,
         `CHM13 count` = chm13_n,
         `CHM13 % of LINEs` = chm13_pct)

kt(t3, caption = paste0(
  "Table 3. LINE family breakdown  ",
  "(hg38 total LINEs n = ", N(.hg38_line_n),
  ";  CHM13 total LINEs n = ", N(.chm13_line_n), ")"
))
```

---

# Liftover Results {#liftover}

## Summary: Both Directions {#both-directions}

**Source:** hg38 → CHM13 row uses `hg38_rmsk.txt.gz`; CHM13 → hg38 row
uses `T2T-CHM13v2.0_RepeatMasker.bed`. Percentages are the fraction of
source loci that failed liftover.

```{r table6}
t6 <- read.delim(file.path(data_dir, "t6_liftover_both.tsv"))
t6_out <- t6 %>%
  mutate(
    mapped_pct   = paste0(round(mapped   / source_total * 100, 1), "%"),
    unmapped_pct = paste0(round(unmapped / source_total * 100, 1), "%")
  ) %>%
  select(Direction = direction,
         `Source loci (n)` = source_total,
         `Mapped` = mapped,
         `Mapped %` = mapped_pct,
         `Unmapped (assembly-specific)` = unmapped,
         `Unmapped %` = unmapped_pct)

kable(t6_out,
      caption = paste0(
        "Table 6. Liftover summary — both directions  ",
        "(hg38 source n = ", N(.hg38_total),
        ";  CHM13 source n = ", N(.chm13_total), ")"
      ),
      format.args = list(big.mark = ","),
      align = c("l", "r", "r", "r", "r", "r")) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "bordered"),
                full_width = FALSE, position = "left") %>%
  row_spec(2, bold = TRUE, color = "white", background = "#8e44ad")
```

> **CHM13 has more total unmapped loci (`r N(.chm13_unmap_all)` vs `r N(.hg38_unmap_all)`)**, but this is almost entirely
> driven by Satellite repeats in centromeres newly resolved by T2T sequencing —
> not by mobile TE insertions. For LINE/SINE elements specifically, hg38 has
> **2.7× more** assembly-specific loci than CHM13 (see Section 5.5).

## hg38 → CHM13: Unmapped by Class {#hg38-unmapped-class}

**Source:** `hg38_rmsk.txt.gz`. The `% of unmapped` column is the fraction of
the `r N(.hg38_unmap_all)` total hg38-specific loci; the `% of hg38 class total`
column is the fraction of all hg38 loci in that class that failed liftover.

```{r table4a}
t4a <- read.delim(file.path(data_dir, "t4a_unmapped_by_class.tsv"))
total_unmap <- sum(t4a$unmapped_n)
t4a <- t4a %>%
  mutate(
    `% of unmapped`        = paste0(round(unmapped_n / total_unmap * 100, 1), "%"),
    `% of hg38 class total`= paste0(round(unmapped_n / hg38_total_in_class * 100, 1), "%")
  ) %>%
  select(Class = class,
         `Unmapped count` = unmapped_n,
         `% of unmapped`,
         `% of hg38 class total`)

kt(t4a, caption = paste0(
  "Table 4a. hg38-specific unmapped elements by class  ",
  "(n = ", N(.hg38_unmap_all), " hg38-specific loci total)"
))
```

## hg38 → CHM13: Top 20 Unmapped Subfamilies {#hg38-top-subfam}

**Source:** `hg38_rmsk.txt.gz`. Percentages are the fraction of all
`r N(.hg38_unmap_all)` hg38-specific unmapped loci.

```{r table4b}
t4b <- read.delim(file.path(data_dir, "t4b_top_subfamilies.tsv"))
total_unmap_all <- sum(t4b$count)
t4b <- t4b %>%
  mutate(`% of hg38-specific unmapped` = paste0(round(count / .hg38_unmap_all * 100, 1), "%")) %>%
  select(Subfamily = subfamily, Count = count, `% of hg38-specific unmapped`)

kable(t4b,
      caption = paste0(
        "Table 4b. Top 20 hg38-specific (unmapped) subfamilies  ",
        "(top 20 of ", N(.hg38_unmap_all), " total hg38-specific loci)"
      ),
      format.args = list(big.mark = ","),
      align = c("l", "r", "r")) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "bordered"),
                full_width = FALSE, position = "left")
```

## CHM13 → hg38: Unmapped by Class {#chm13-unmapped-class}

**Source:** `T2T-CHM13v2.0_RepeatMasker.bed`. The `% of unmapped` column is
the fraction of the `r N(.chm13_unmap_all)` total CHM13-specific loci;
the `% of CHM13 class total` column is the fraction of all CHM13 loci in
that class that failed liftover. Note that **Satellite** and **Simple_repeat**
classes dominate — centromeric repeats that T2T resolved for the first time
but that exist in hg38 only as collapsed or absent sequence.

```{r table7}
t7 <- read.delim(file.path(data_dir, "t7_chm13_unmapped_by_class.tsv"))
total_unmap7 <- sum(t7$unmapped_n)
t7_out <- t7 %>%
  mutate(
    `% of unmapped`          = paste0(round(unmapped_n / total_unmap7 * 100, 1), "%"),
    `% of CHM13 class total` = paste0(round(unmapped_n / chm13_total_in_class * 100, 1), "%")
  ) %>%
  select(Class = class,
         `Unmapped count` = unmapped_n,
         `% of unmapped`,
         `% of CHM13 class total`)

kable(t7_out,
      caption = paste0(
        "Table 7. CHM13-specific unmapped elements by class  ",
        "(n = ", N(.chm13_unmap_all), " CHM13-specific loci total)"
      ),
      format.args = list(big.mark = ","),
      align = c("l", "r", "r", "r")) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "bordered"),
                full_width = FALSE, position = "left") %>%
  row_spec(which(t7_out$Class == "Satellite"), bold = TRUE, color = "white", background = "#c0392b") %>%
  row_spec(nrow(t7_out), bold = TRUE, background = "#f0f0f0")
```

## CHM13 → hg38: Top 20 Unmapped Subfamilies {#chm13-top-subfam}

**Source:** `T2T-CHM13v2.0_RepeatMasker.bed`. Percentages are the fraction
of all `r N(.chm13_unmap_all)` CHM13-specific unmapped loci.

```{r table8}
t8 <- read.delim(file.path(data_dir, "t8_chm13_top_subfamilies.tsv"))
t8_out <- t8 %>%
  mutate(`% of CHM13-specific unmapped` = paste0(round(count / .chm13_unmap_all * 100, 1), "%")) %>%
  select(Subfamily = subfamily, Count = count, `% of CHM13-specific unmapped`)

kable(t8_out,
      caption = paste0(
        "Table 8. Top 20 CHM13-specific (unmapped) subfamilies  ",
        "(top 20 of ", N(.chm13_unmap_all), " total CHM13-specific loci)"
      ),
      format.args = list(big.mark = ","),
      align = c("l", "r", "r")) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "bordered"),
                full_width = FALSE, position = "left")
```

> **HSATII and (ATCGAATGGA)n** account for the majority of CHM13-specific elements —
> these are human satellite II repeats concentrated in pericentromeric regions,
> which T2T-CHM13 resolved completely for the first time. Their "absence" from hg38
> is an assembly artifact, not a biological insertion difference.

## LINE/SINE-Specific Comparison {#linesine-comparison}

**Source:** LINE/SINE loci from `hg38_rmsk.txt.gz` (hg38 row) and from
`T2T-CHM13v2.0_RepeatMasker.bed` (CHM13 row). Percentages are the fraction
of source LINE/SINE loci that failed liftover. Restricting to LINE and SINE
removes the satellite/centromere signal and isolates the biologically relevant
TE-insertion comparison.

```{r table9}
t9 <- read.delim(file.path(data_dir, "t9_linesine_comparison.tsv"))
t9_out <- t9 %>%
  mutate(
    unmapped_pct_fmt = paste0(round(unmapped_pct, 2), "%"),
    fold = c(round(t9$unmapped_pct[1] / t9$unmapped_pct[2], 1), 1.0)
  ) %>%
  select(Direction = direction,
         `Source LINE/SINE loci (n)` = source_linesine_total,
         `Unmapped (assembly-specific)` = unmapped_linesine,
         `% of source LINE/SINE` = unmapped_pct_fmt,
         `Fold vs CHM13` = fold)

kable(t9_out,
      caption = paste0(
        "Table 9. LINE/SINE-only assembly-specific comparison  ",
        "(hg38 LINE/SINE n = ", N(.hg38_ls_total),
        ";  CHM13 LINE/SINE n = ", N(.chm13_ls_total), ")"
      ),
      format.args = list(big.mark = ","),
      align = c("l", "r", "r", "r", "r")) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "bordered"),
                full_width = FALSE, position = "left") %>%
  row_spec(1, bold = TRUE, color = "white", background = "#27ae60") %>%
  row_spec(nrow(t9_out), bold = TRUE, background = "#f0f0f0")
```

> **Among LINE and SINE elements, hg38 has 2.7× more assembly-specific loci than CHM13**
> (0.96% vs 0.36% unmapped rate). This is the primary TE-biology signal: hg38 contains
> more polymorphic or unresolved TE insertions than the T2T reference genome.

---

# Young vs. Old Alu — The Key Finding {#alu-age}

**Source:** hg38 background Alu counts from `hg38_rmsk.txt.gz`; unmapped Alu
sets from the liftover outputs (hg38-specific from hg38→CHM13; CHM13-specific
from CHM13→hg38). Percentages within each column are independent: each column
sums to 100% over its own total.

Among Alu elements that failed liftover, **young AluY subfamilies are massively
over-represented** in the CHM13-specific set relative to the hg38 background.
This is consistent with young Alu elements still retrotransposing and creating
individual-specific insertions that are absent from other assemblies.

```{r table5}
t5 <- read.delim(file.path(data_dir, "t5_alu_age.tsv"))

totals <- t5[t5$age_group == "Total", ]
t5_display <- t5 %>% filter(age_group != "Total") %>%
  mutate(
    hg38_bg_pct     = paste0(round(hg38_bg_n    / totals$hg38_bg_n    * 100, 1), "%"),
    hg38_unmap_pct  = paste0(round(hg38_unmap_n / totals$hg38_unmap_n * 100, 1), "%"),
    chm13_unmap_pct = paste0(round(chm13_unmap_n/ totals$chm13_unmap_n* 100, 1), "%")
  )

total_row <- data.frame(
  age_group       = "Total",
  hg38_bg_n       = totals$hg38_bg_n,
  hg38_unmap_n    = totals$hg38_unmap_n,
  chm13_unmap_n   = totals$chm13_unmap_n,
  hg38_bg_pct     = "100%",
  hg38_unmap_pct  = "100%",
  chm13_unmap_pct = "100%"
)
t5_display <- bind_rows(t5_display, total_row)

t5_out <- t5_display %>%
  select(
    `Alu age group`                                         = age_group,
    `hg38 background\nn = 1,228,449`                       = hg38_bg_n,
    ` `                                                     = hg38_bg_pct,
    `hg38-specific unmapped\nn = 14,531`                   = hg38_unmap_n,
    `  `                                                    = hg38_unmap_pct,
    `CHM13-specific unmapped\nn = 10,068`                  = chm13_unmap_n,
    `   `                                                   = chm13_unmap_pct
  )

kable(t5_out,
      caption = paste0(
        "Table 5. Young vs. old Alu — assembly-specific unmapped vs. hg38 background  ",
        "(hg38 Alu background n = ", N(totals$hg38_bg_n),
        ";  hg38-specific Alus n = ", N(totals$hg38_unmap_n),
        ";  CHM13-specific Alus n = ", N(totals$chm13_unmap_n), ")"
      ),
      format.args = list(big.mark = ","),
      align = c("l", "r", "r", "r", "r", "r", "r")) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "bordered"),
                full_width = FALSE, position = "left") %>%
  add_header_above(c(" " = 1,
                     "hg38 background" = 2,
                     "hg38-specific unmapped" = 2,
                     "CHM13-specific unmapped" = 2)) %>%
  row_spec(which(t5_out[[1]] == "Young (AluY*)"),
           bold = TRUE, color = "white", background = "#2980b9") %>%
  row_spec(nrow(t5_out), bold = TRUE, background = "#f0f0f0") %>%
  footnote(general = paste0(
    "AluY* = young Alu subfamilies (~1-5 Mya, still occasionally retrotransposing). ",
    "AluS* = intermediate (~5-20 Mya). AluJ* = oldest (~20-65 Mya, universally fixed). ",
    "CHM13-specific unmapped set derived from CHM13→hg38 liftover."
  ))
```

> **Interpretation:** Young AluY* elements make up ~12% of all hg38 Alu loci (background)
> but **76% of CHM13-specific unmapped Alu loci**. Young Alus are the elements
> most likely to be absent from the other assembly — direct evidence that recently-mobile
> TEs are still creating individual-specific (polymorphic) insertions that lack
> fixed genomic addresses across the human population.

---

# Methods {#methods}

| Item | Details |
|------|---------|
| hg38 RepeatMasker | UCSC `rmsk.txt.gz`, downloaded October 2022 |
| CHM13 RepeatMasker | T2T-CHM13v2.0 RepeatMasker BED (project `references/`) |
| Liftover tool | CrossMap v0.7.0 |
| Chain files | UCSC `hg38ToHs1.over.chain.gz` and `hs1ToHg38.over.chain.gz` |
| Architecture | aarch64 Linux (UCSC liftOver binary not available; CrossMap used instead) |
| Reference | Hoyt et al. 2022, *Science* 376, eabk3084 (reported 20,427 CHM13-specific TEs) |

**Cleanup:** to remove all downloaded data and results: `rm -rf analysis/liftover/data/ analysis/liftover/results/`
