---
title: "hg38 ↔ T2T-CHM13 Transposable Element Liftover Analysis"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    theme: flatly
    toc: true
    toc_float: true
    toc_depth: 3
    number_sections: true
    df_print: paged
  pdf_document:
    toc: true
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
library(knitr)
library(kableExtra)
library(dplyr)

# Data lives in results/ relative to this Rmd file
data_dir <- file.path(dirname(knitr::current_input(dir = TRUE)), "results")

kt <- function(df, caption = NULL, ...) {
  kable(df, caption = caption, format.args = list(big.mark = ","),
        align = c("l", rep("r", ncol(df) - 1)), ...) %>%
    kable_styling(bootstrap_options = c("striped", "hover", "condensed", "bordered"),
                  full_width = FALSE, position = "left") %>%
    row_spec(nrow(df), bold = TRUE, background = "#f0f0f0")
}
```

> **Motivation:** Are there more assembly-specific TE loci in hg38 than in T2T-CHM13?
> And are young, recently-mobile Alu elements disproportionately absent from
> the other assembly — evidence that they lack stable genomic addresses across individuals?

---

# Total RepeatMasker Annotations {#total}

hg38 and T2T-CHM13v2.0 RepeatMasker annotations across all repeat classes.

```{r table1}
t1 <- read.delim(file.path(data_dir, "t1_class_counts.tsv"))
t1 <- t1 %>%
  mutate(
    hg38_pct  = ifelse(class == "Total", "100%", paste0(round(hg38_n  / hg38_n[class=="Total"]  * 100, 1), "%")),
    chm13_pct = ifelse(class == "Total", "100%", paste0(round(chm13_n / chm13_n[class=="Total"] * 100, 1), "%"))
  ) %>%
  select(Class = class,
         `hg38 count` = hg38_n,
         `hg38 %` = hg38_pct,
         `CHM13 count` = chm13_n,
         `CHM13 %` = chm13_pct)

kt(t1, caption = "Table 1. Total TE annotations by class")
```

---

# SINE and LINE Family Breakdown {#families}

## SINE Families

```{r table2}
t2 <- read.delim(file.path(data_dir, "t2_sine_families.tsv"))
total_row <- t2[t2$family == "Total", ]
t2 <- t2 %>%
  mutate(
    hg38_pct  = ifelse(family == "Total", "100%",
                  paste0(round(hg38_n  / total_row$hg38_n  * 100, 1), "%")),
    chm13_pct = ifelse(family == "Total", "100%",
                  paste0(round(chm13_n / total_row$chm13_n * 100, 1), "%"))
  ) %>%
  select(Family = family,
         `hg38 count` = hg38_n,
         `hg38 %` = hg38_pct,
         `CHM13 count` = chm13_n,
         `CHM13 %` = chm13_pct)

kt(t2, caption = "Table 2. SINE family breakdown")
```

## LINE Families

```{r table3}
t3 <- read.delim(file.path(data_dir, "t3_line_families.tsv"))
total_row3 <- t3[t3$family == "Total", ]
t3 <- t3 %>%
  mutate(
    hg38_pct  = ifelse(family == "Total", "100%",
                  paste0(round(hg38_n  / total_row3$hg38_n  * 100, 1), "%")),
    chm13_pct = ifelse(family == "Total", "100%",
                  paste0(round(chm13_n / total_row3$chm13_n * 100, 1), "%"))
  ) %>%
  select(Family = family,
         `hg38 count` = hg38_n,
         `hg38 %` = hg38_pct,
         `CHM13 count` = chm13_n,
         `CHM13 %` = chm13_pct)

kt(t3, caption = "Table 3. LINE family breakdown")
```

---

# hg38 → CHM13 Liftover Results {#liftover}

All **5,683,690** hg38 RepeatMasker loci were lifted to T2T-CHM13 coordinates
using CrossMap v0.7.0 with the UCSC hg38→hs1 chain file.

## Summary

```{r table4}
t4 <- read.delim(file.path(data_dir, "t4_liftover_summary.tsv"))
total_n <- t4$count[t4$outcome == "Total"]
t4 <- t4 %>%
  mutate(pct = ifelse(outcome == "Total", "100%",
                paste0(round(count / total_n * 100, 1), "%"))) %>%
  select(Outcome = outcome, Count = count, `% of hg38 total` = pct)

kable(t4, caption = "Table 4. Liftover summary",
      format.args = list(big.mark = ","),
      align = c("l", "r", "r")) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "bordered"),
                full_width = FALSE, position = "left") %>%
  row_spec(which(t4$Outcome == "Unmapped (hg38-specific)"),
           bold = TRUE, color = "white", background = "#c0392b") %>%
  row_spec(nrow(t4), bold = TRUE, background = "#f0f0f0")
```

> **58,933 hg38 loci have no equivalent position in CHM13.** These represent a mix of:
> hg38 assembly errors, collapsed repeats resolved differently in T2T, and genuine
> polymorphic insertions absent from CHM13 (this individual's genome).

## Unmapped Elements by Class

```{r table4a}
t4a <- read.delim(file.path(data_dir, "t4a_unmapped_by_class.tsv"))
total_unmap <- sum(t4a$unmapped_n)
t4a <- t4a %>%
  mutate(
    `% of unmapped`        = paste0(round(unmapped_n / total_unmap * 100, 1), "%"),
    `% of hg38 class total`= paste0(round(unmapped_n / hg38_total_in_class * 100, 1), "%")
  ) %>%
  select(Class = class,
         `Unmapped count` = unmapped_n,
         `% of unmapped`,
         `% of hg38 class total`)

kt(t4a, caption = "Table 4a. Unmapped elements by class")
```

## Top 20 Unmapped Subfamilies

```{r table4b}
t4b <- read.delim(file.path(data_dir, "t4b_top_subfamilies.tsv"))
total_unmap_all <- sum(t4b$count)
t4b <- t4b %>%
  mutate(`% of unmapped` = paste0(round(count / total_unmap_all * 100, 1), "%")) %>%
  select(Subfamily = subfamily, Count = count, `% of unmapped`)

kable(t4b, caption = "Table 4b. Top 20 hg38-specific (unmapped) subfamilies",
      format.args = list(big.mark = ","),
      align = c("l", "r", "r")) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "bordered"),
                full_width = FALSE, position = "left")
```

---

# Young vs. Old Alu — The Key Finding {#alu-age}

Among Alu elements that failed liftover, **young AluY subfamilies are massively
over-represented** in the CHM13-specific set relative to the hg38 background.
This is consistent with young Alu elements still retrotransposing and creating
individual-specific insertions that are absent from other assemblies.

```{r table5}
t5 <- read.delim(file.path(data_dir, "t5_alu_age.tsv"))

totals <- t5[t5$age_group == "Total", ]
t5_display <- t5 %>% filter(age_group != "Total") %>%
  mutate(
    hg38_bg_pct     = paste0(round(hg38_bg_n    / totals$hg38_bg_n    * 100, 1), "%"),
    hg38_unmap_pct  = paste0(round(hg38_unmap_n / totals$hg38_unmap_n * 100, 1), "%"),
    chm13_unmap_pct = paste0(round(chm13_unmap_n/ totals$chm13_unmap_n* 100, 1), "%")
  )

# Add total row
total_row <- data.frame(
  age_group       = "Total",
  hg38_bg_n       = totals$hg38_bg_n,
  hg38_unmap_n    = totals$hg38_unmap_n,
  chm13_unmap_n   = totals$chm13_unmap_n,
  hg38_bg_pct     = "100%",
  hg38_unmap_pct  = "100%",
  chm13_unmap_pct = "100%"
)
t5_display <- bind_rows(t5_display, total_row)

t5_out <- t5_display %>%
  select(
    `Alu age group`                              = age_group,
    `hg38 background\nn = 1,228,449`            = hg38_bg_n,
    ` `                                          = hg38_bg_pct,
    `hg38-specific unmapped\nn = 14,531`         = hg38_unmap_n,
    `  `                                         = hg38_unmap_pct,
    `CHM13-specific unmapped\nn = 10,068`        = chm13_unmap_n,
    `   `                                        = chm13_unmap_pct
  )

kable(t5_out,
      caption = "Table 5. Young vs. old Alu in assembly-specific unmapped sets vs. hg38 background",
      format.args = list(big.mark = ","),
      align = c("l", "r", "r", "r", "r", "r", "r")) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "bordered"),
                full_width = FALSE, position = "left") %>%
  add_header_above(c(" " = 1,
                     "hg38 background" = 2,
                     "hg38-specific unmapped" = 2,
                     "CHM13-specific unmapped" = 2)) %>%
  row_spec(which(t5_out[[1]] == "Young (AluY*)"),
           bold = TRUE, color = "white", background = "#2980b9") %>%
  row_spec(nrow(t5_out), bold = TRUE, background = "#f0f0f0") %>%
  footnote(general = paste0(
    "AluY* = young Alu subfamilies (~1-5 Mya, still occasionally retrotransposing). ",
    "AluS* = intermediate (~5-20 Mya). AluJ* = oldest (~20-65 Mya, universally fixed). ",
    "CHM13-specific unmapped from CHM13→hg38 liftover."
  ))
```

> **Interpretation:** Young AluY* elements make up ~12% of all hg38 Alu loci (background)
> but **76% of CHM13-specific unmapped Alu loci**. Young Alus are the elements
> most likely to be absent from the other assembly — direct evidence that recently-mobile
> TEs are still creating individual-specific (polymorphic) insertions that lack
> fixed genomic addresses across the human population.

---

# Methods {#methods}

| Item | Details |
|------|---------|
| hg38 RepeatMasker | UCSC `rmsk.txt.gz`, downloaded October 2022 |
| CHM13 RepeatMasker | T2T-CHM13v2.0 RepeatMasker BED (project `references/`) |
| Liftover tool | CrossMap v0.7.0 |
| Chain files | UCSC hg38→hs1 and hs1→hg38 |
| Platform | aarch64 Linux (UCSC liftOver binary not available; CrossMap used instead) |
| Reference | Hoyt et al. 2022, *Science* 376, eabk3084 (reported 20,427 CHM13-specific TEs across all classes) |

**Cleanup:** to remove all downloaded data and results: `rm -rf analysis/liftover/data/ analysis/liftover/results/`
