---
title: "T2T vs hg38 for TE Quantification: Patient Data vs. Assembly Architecture"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    theme: flatly
    toc: true
    toc_float: true
    toc_depth: 3
    number_sections: true
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
library(knitr)
library(kableExtra)
library(dplyr)

liftover_dir  <- file.path(dirname(knitr::current_input(dir = TRUE)), "results")
patient_dir   <- file.path(liftover_dir, "patient_data")

N <- function(n) format(as.numeric(n), big.mark = ",", scientific = FALSE)

kt <- function(df, caption = NULL, ...) {
  kable(df, caption = caption, format.args = list(big.mark = ","),
        align = c("l", rep("r", ncol(df) - 1)), ...) %>%
    kable_styling(bootstrap_options = c("striped", "hover", "condensed", "bordered"),
                  full_width = FALSE, position = "left") %>%
    row_spec(nrow(df), bold = TRUE, background = "#f0f0f0")
}
```

> **Question:** A previous analysis of a patient snRNAseq dataset (Kumar Sample-17)
> found that T2T-CHM13v2.0 detected more TEs than hg38 when running soloTE.
> A liftover analysis of RepeatMasker annotations, however, shows that hg38
> has *more* assembly-specific LINE/SINE loci than T2T. Are these findings
> contradictory? This report reconciles them.

---

# What We Are Comparing {#overview}

| | soloTEmvp (patient data) | soloTEmvpWGsim (liftover) |
|---|---|---|
| **Data type** | Real snRNAseq — Kumar Sample-17, 3,796 cells | RepeatMasker BED annotations — no reads |
| **Question asked** | How many TE loci/counts does each reference detect? | How many annotated TE loci exist in one assembly but not the other? |
| **Key metric** | soloTE read counts per cell | % of loci that fail liftover to the other assembly |
| **Unit** | Detected locus × cell matrix entries | Genomic coordinate intervals |

These measure fundamentally different things. A locus can be annotated in hg38 but never
generate a sequencing read (inaccessible, in a gap, or in a polymorphic region absent from
most patients). Conversely, T2T's better assembly can let reads align *more uniquely*
to loci that would be forced into multi-mapping in hg38.

---

# Patient Data Results — Fair Comparison {#patient}

## The Aligner Confound {#confound}

The patient data analysis went through two rounds of correction. The initial comparison
was confounded by a difference in aligners:

```{r confound-table}
conf <- data.frame(
  ` ` = c("Aligner", "--outFilterMultimapNmax", "TEs mapping to 20–100 loci", "Result"),
  `Previous confounded` = c("T2T: STARsolo / hg38: CellRanger",
                             "T2T: 100 / hg38: 20 (hardcoded)",
                             "Kept (T2T) vs discarded (hg38)",
                             "3.5× counts/cell — CONFOUNDED"),
  `Fair comparison (master)` = c("Both STARsolo",
                                  "Both: 100 (matched)",
                                  "Kept in both",
                                  "1.15× counts/cell — true reference effect"),
  check.names = FALSE
)

kable(conf, align = c("l", "r", "r")) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "bordered"),
                full_width = FALSE, position = "left") %>%
  row_spec(4, bold = TRUE) %>%
  row_spec(2, color = "#c0392b") %>%
  row_spec(3, color = "#c0392b") %>%
  column_spec(3, background = "#eafaf1")
```

TEs are highly repetitive and frequently multimap to 20–100 genomic loci. CellRanger's
hardcoded limit of 20 systematically discards those reads, disproportionately penalising
TE detection. When both samples are aligned with STARsolo at the same limit of 100, the
3.5× inflation collapses to a **real ~15% T2T advantage**.

## Feature Counts at Each Matrix Level {#features}

soloTE produces five output levels: locus, subfamily, family, class, and legacy.
The fair comparison results (master branch, April 21 2026) are:

```{r feature-counts}
fc <- read.csv(file.path(patient_dir, "fair_feature_counts.csv"))
fc_out <- fc %>%
  mutate(
    fold_fmt = round(fold_change, 3),
    direction = ifelse(fold_change >= 1, "T2T > hg38", "hg38 > T2T")
  ) %>%
  select(
    Level = level,
    `T2T features` = t2t_features,
    `hg38 features` = hg38_features,
    `Fold (T2T/hg38)` = fold_fmt,
    Direction = direction
  )

kable(fc_out, caption = "Table P1. soloTE feature counts at each matrix level (fair comparison, n = 3,796 cells)",
      format.args = list(big.mark = ","),
      align = c("l", "r", "r", "r", "l")) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "bordered"),
                full_width = FALSE, position = "left") %>%
  row_spec(which(fc_out$`Fold (T2T/hg38)` > 1), background = "#eafaf1") %>%
  row_spec(which(fc_out$`Fold (T2T/hg38)` < 1), background = "#fdf2f8")
```

> **Locus level (locustes):** T2T has **1.16× more features** — the primary signal.
>
> **Subfamily/family/class levels:** hg38 has **more features** (fold < 1). This is
> expected: in T2T, better assembly resolves repetitive regions so more reads map
> *uniquely* and land at locus level. In hg38, more reads multi-map and are pushed
> up to subfamily/family level, creating more features at those levels but with
> lower confidence assignment.

## TE Detection by Threshold {#detection}

Not all annotated loci are robustly expressed. The table below shows loci meeting
various detection thresholds. The T2T advantage (~1.15–1.16×) is **consistent across
all thresholds**, confirming it is a real reference-quality effect.

```{r locus-detection}
ld <- read.csv(file.path(patient_dir, "fair_locus_detection.csv"), check.names = FALSE)
ld_out <- ld %>%
  mutate(
    `T2T %` = paste0(round(T2T_percent, 1), "%"),
    `hg38 %` = paste0(round(hg38_percent, 1), "%"),
    fold_fmt = round(fold_change, 3)
  ) %>%
  select(
    Threshold = threshold,
    `T2T loci` = T2T_loci,
    `T2T %` ,
    `hg38 loci` = hg38_loci,
    `hg38 %`,
    `Fold (T2T/hg38)` = fold_fmt
  )

kable(ld_out,
      caption = "Table P2. Loci detected at each threshold (n = 3,796 cells, locus level only)",
      format.args = list(big.mark = ","),
      align = c("l", "r", "r", "r", "r", "r")) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "bordered"),
                full_width = FALSE, position = "left") %>%
  row_spec(1, bold = TRUE) %>%
  row_spec(which(ld_out$Threshold == "≥10 cells AND ≥50 counts"),
           bold = TRUE, background = "#d6eaf8")
```

> Only **~29% of annotated loci** are robustly detected (≥10 cells). The remaining 71%
> have trace counts from multi-mapper spreading and should not be reported as "expressed."
> The ≥10 cells AND ≥50 counts threshold (highlighted) is the recommended threshold
> for downstream analyses.

## Per-Cell TE Count Summary {#per-cell}

```{r per-cell}
pc <- data.frame(
  Metric = c("Total TE counts", "Mean counts / cell", "Median counts / cell", "Cells"),
  T2T    = c("37,343,116", "9,837.5", "6,150.5", "3,796"),
  hg38   = c("32,431,222", "8,543.5", "5,305.5", "3,796"),
  `Fold (T2T/hg38)` = c("1.15×", "1.15×", "1.16×", "—"),
  check.names = FALSE
)
kt(pc, caption = "Table P3. Per-cell TE count summary (fair comparison)")
```

## Top TE Families {#families}

```{r top-families}
fe <- read.csv(file.path(patient_dir, "fair_family_expression.csv"))
fe_top <- fe %>%
  arrange(desc(t2t_counts)) %>%
  head(15) %>%
  mutate(
    family  = sub("SoloTE\\|", "", family),
    log2fc  = round(log2fc, 3),
    flag    = ifelse(log2fc > 0.5, "T2T enriched", ifelse(log2fc < -0.5, "hg38 enriched", "similar"))
  ) %>%
  select(
    Family        = family,
    `T2T counts`  = t2t_counts,
    `hg38 counts` = hg38_counts,
    `log2FC (T2T/hg38)` = log2fc,
    ` ` = flag
  )

kable(fe_top,
      caption = "Table P4. Top 15 TE families by T2T expression (fair comparison, n = 30 shared families)",
      format.args = list(big.mark = ","),
      align = c("l", "r", "r", "r", "l")) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "bordered"),
                full_width = FALSE, position = "left") %>%
  row_spec(which(fe_top$`log2FC (T2T/hg38)` > 0.5),
           background = "#eafaf1")
```

## Assembly-Specific Families {#assembly-families}

```{r exclusive-families}
t2t_only <- readLines(file.path(patient_dir, "fair_t2t_only_families.txt"))
hg38_only <- readLines(file.path(patient_dir, "fair_hg38_only_families.txt"))
t2t_only <- sub("SoloTE\\|", "", t2t_only[t2t_only != ""])
hg38_only <- sub("SoloTE\\|", "", hg38_only[hg38_only != ""])

excl <- data.frame(
  `T2T-only families (n = 12)` = c(t2t_only, rep("", max(0, length(hg38_only) - length(t2t_only)))),
  `hg38-only families (n = 3)` = c(hg38_only, rep("", max(0, length(t2t_only) - length(hg38_only)))),
  check.names = FALSE
)

kable(excl,
      caption = "Table P5. TE families detected in only one reference (fair comparison)") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "bordered"),
                full_width = FALSE, position = "left") %>%
  column_spec(1, background = "#eafaf1") %>%
  column_spec(2, background = "#fdf2f8")
```

---

# Assembly Architecture — Liftover Analysis {#liftover}

The liftover analysis compares RepeatMasker *annotations* directly between hg38 and T2T,
independent of any reads. Source files:

- **hg38**: `hg38_rmsk.txt.gz` (UCSC, Oct 2022) — 5,683,690 total annotations
- **T2T**: `T2T-CHM13v2.0_RepeatMasker.bed` (project `references/`) — 5,590,282 total annotations

## Total Annotations and LINE/SINE Counts {#annotations}

```{r total-annotations}
t1 <- read.delim(file.path(liftover_dir, "t1_class_counts.tsv"))
t9 <- read.delim(file.path(liftover_dir, "t9_linesine_comparison.tsv"))

hg38_ls <- t9$source_linesine_total[1]
chm13_ls <- t9$source_linesine_total[2]

annot <- data.frame(
  ` ` = c("All classes", "LINE + SINE only"),
  `hg38 (n)` = c(
    format(t1$hg38_n[t1$class == "Total"], big.mark = ","),
    format(hg38_ls, big.mark = ",")
  ),
  `T2T (n)` = c(
    format(t1$chm13_n[t1$class == "Total"], big.mark = ","),
    format(chm13_ls, big.mark = ",")
  ),
  `T2T vs hg38` = c(
    paste0(round(t1$chm13_n[t1$class=="Total"] / t1$hg38_n[t1$class=="Total"], 3), "×"),
    paste0(round(chm13_ls / hg38_ls, 3), "×")
  ),
  check.names = FALSE
)

kable(annot,
      caption = "Table L1. Total RepeatMasker annotations — T2T vs hg38",
      align = c("l", "r", "r", "r")) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "bordered"),
                full_width = FALSE, position = "left") %>%
  row_spec(2, bold = TRUE, background = "#eafaf1")
```

> T2T has **2.7% more LINE/SINE loci** than hg38 in total annotation count
> (3,618,644 vs 3,525,112). This is a modest but consistent difference.

## Assembly-Specific Loci — Liftover Results {#assembly-specific}

When each assembly's annotations are lifted to the other assembly's coordinate space,
loci that fail = assembly-specific (present in source but absent/unresolvable in target).

```{r assembly-specific}
t9_out <- t9 %>%
  mutate(
    unmapped_pct_fmt = paste0(round(unmapped_pct, 2), "%"),
    fold = c(round(t9$unmapped_pct[1] / t9$unmapped_pct[2], 1), 1.0)
  ) %>%
  select(
    Direction = direction,
    `Source LINE/SINE loci` = source_linesine_total,
    `Assembly-specific (unmapped)` = unmapped_linesine,
    `% of source` = unmapped_pct_fmt,
    `Fold vs T2T` = fold
  )

kable(t9_out,
      caption = paste0("Table L2. Assembly-specific LINE/SINE loci — both liftover directions",
                       "  (hg38 LINE/SINE n = ", N(hg38_ls), ";  T2T LINE/SINE n = ", N(chm13_ls), ")"),
      format.args = list(big.mark = ","),
      align = c("l", "r", "r", "r", "r")) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "bordered"),
                full_width = FALSE, position = "left") %>%
  row_spec(1, bold = TRUE, color = "white", background = "#e74c3c") %>%
  row_spec(2, bold = TRUE, color = "white", background = "#27ae60")
```

> **hg38 has 2.7× more assembly-specific LINE/SINE loci** (36,480 vs 13,155).
> This means hg38 annotates nearly three times as many LINE/SINE loci that have no
> counterpart in T2T. These appear to be in regions that T2T resolved differently,
> or genuine polymorphic insertions absent from the CHM13 cell line.

## Young Alu Enrichment {#alu}

```{r alu-summary}
t5 <- read.delim(file.path(liftover_dir, "t5_alu_age.tsv"))
totals <- t5[t5$age_group == "Total", ]
t5_s <- t5 %>%
  filter(age_group != "Total") %>%
  mutate(
    hg38_bg_pct     = paste0(round(hg38_bg_n    / totals$hg38_bg_n    * 100, 1), "%"),
    chm13_unmap_pct = paste0(round(chm13_unmap_n / totals$chm13_unmap_n * 100, 1), "%")
  ) %>%
  select(
    `Alu age group`              = age_group,
    `hg38 background`            = hg38_bg_n,
    `hg38 bg %`                  = hg38_bg_pct,
    `T2T-specific unmapped`      = chm13_unmap_n,
    `T2T-specific %`             = chm13_unmap_pct
  )
# add total row
t5_s <- bind_rows(t5_s, data.frame(
  `Alu age group` = "Total",
  `hg38 background` = totals$hg38_bg_n,
  `hg38 bg %` = "100%",
  `T2T-specific unmapped` = totals$chm13_unmap_n,
  `T2T-specific %` = "100%",
  check.names = FALSE
))

kable(t5_s,
      caption = paste0("Table L3. Alu age composition: hg38 background vs T2T-specific unmapped",
                       "  (hg38 Alu bg n = ", N(totals$hg38_bg_n), ";  T2T-specific Alus n = ",
                       N(totals$chm13_unmap_n), ")"),
      format.args = list(big.mark = ","),
      align = c("l", "r", "r", "r", "r")) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "bordered"),
                full_width = FALSE, position = "left") %>%
  row_spec(which(t5_s[[1]] == "Young (AluY*)"),
           bold = TRUE, color = "white", background = "#2980b9") %>%
  row_spec(nrow(t5_s), bold = TRUE, background = "#f0f0f0") %>%
  footnote(general = "AluY* ≈ young (~1–5 Mya, still mobile). AluS* ≈ intermediate (~5–20 Mya). AluJ* ≈ old (~20–65 Mya, fixed).")
```

> **76% of T2T-specific unmapped Alus are young AluY*** (vs 12% of all hg38 Alus).
> These are polymorphic insertions unique to the CHM13 individual that have no hg38
> counterpart — loci that T2T adds over hg38 and that are candidates for
> individual-specific TE expression.

---

# Reconciliation {#reconciliation}

The two analyses appear contradictory at first glance:

- **Liftover**: hg38 has 2.7× more assembly-specific LINE/SINE loci → hg38 "has more TEs"
- **Patient data**: T2T detects 1.16× more TE loci in scRNAseq → T2T "finds more TEs"

They are compatible because they measure different populations.

## Why hg38 Has More Assembly-Specific Annotations

```{r reconcile-table}
rec <- data.frame(
  `hg38-specific TEs (n = 36,480 LINE/SINE)` = c(
    "In regions with gaps, collapses, or mis-assemblies in hg38",
    "Many are RepeatMasker artifacts adjacent to assembly gaps",
    "Reads from patients rarely map there (inaccessible sequence)",
    "Higher proportion of old/fixed elements (AluS*, AluJ*)",
    "Contribute little to actual scRNAseq TE counts"
  ),
  `T2T-specific TEs (n = 13,155 LINE/SINE)` = c(
    "In regions newly resolved by T2T (gap-filled, centromeric flanks)",
    "Genuine sequence present in CHM13 but absent/collapsed in hg38",
    "76% are young AluY* — polymorphic, potentially active insertions",
    "Can generate reads in T2T-aligned data",
    "Contribute to the ~1.15× T2T advantage in scRNAseq"
  ),
  check.names = FALSE
)

kable(rec, caption = "Table R1. Character of assembly-specific TEs in each direction",
      align = c("l", "l")) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "bordered"),
                full_width = TRUE, position = "left") %>%
  column_spec(1, background = "#fdedec", width = "44%") %>%
  column_spec(2, background = "#eafaf1", width = "44%")
```

## Decomposing the T2T Advantage in Practice

```{r decompose-advantage}
decomp <- data.frame(
  Source = c(
    "T2T has more total LINE/SINE annotations",
    "T2T resolves repetitive regions → fewer false multimappers",
    "T2T-specific polymorphic Alus (young AluY*) accessible to reads",
    "hg38-specific TEs in inaccessible / gap-adjacent regions",
    "Aligner confound (CellRanger 20 vs STARsolo 100 multimapper limit)",
    "Net observed T2T advantage (fair comparison)",
    "Net observed T2T advantage (confounded comparison)"
  ),
  `Loci delta` = c("+93,532 more T2T LINE/SINE loci (+2.7%)",
                   "shifts counts from subfamily → locus level",
                   "+~13k T2T-unique loci with expression potential",
                   "−36k hg38 loci with little expression impact",
                   "not applicable to loci count",
                   "1.16× loci, 1.15× counts/cell",
                   "1.44× loci, 3.5× counts/cell"),
  Status = c("✓ assembly-level", "✓ mapping quality", "✓ biology",
             "✓ biology/assembly", "⚠ confound (fixed)", "✓ true effect", "⚠ inflated"),
  check.names = FALSE
)

kable(decomp, caption = "Table R2. Sources of T2T advantage — annotation vs detection vs confound",
      align = c("l", "l", "l")) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "bordered"),
                full_width = TRUE, position = "left") %>%
  row_spec(which(decomp$Status == "⚠ confound (fixed)"),
           color = "#c0392b") %>%
  row_spec(which(decomp$Status == "⚠ inflated"),
           color = "#c0392b") %>%
  row_spec(which(grepl("Net observed.*fair", decomp$Source)),
           bold = TRUE, background = "#eafaf1") %>%
  row_spec(which(grepl("Net observed.*confound", decomp$Source)),
           bold = TRUE, background = "#fdedec")
```

## Summary: Both Findings Are Correct

```{r summary-box}
summ <- data.frame(
  Finding = c(
    "hg38 has 2.7× more assembly-specific LINE/SINE TEs",
    "T2T detects 1.16× more TE loci in patient scRNAseq",
    "T2T gives 3.5× more counts/cell in original comparison"
  ),
  Correct = c("Yes", "Yes", "Partially"),
  Explanation = c(
    "These are mostly in poorly assembled / gap-adjacent regions that do not generate detectable reads",
    "True reference-quality effect: T2T has more total LINE/SINE loci, better assembly reduces false multimappers, and young AluY insertions in T2T are accessible",
    "1.15× is the true reference effect; the remaining ~3× was aligner confound (CellRanger 20 vs STARsolo 100 multimapper limit)"
  )
)

kable(summ, caption = "Table R3. Reconciliation summary",
      align = c("l", "l", "l")) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "bordered"),
                full_width = TRUE, position = "left") %>%
  column_spec(2, width = "6em",
              bold = TRUE,
              color = ifelse(summ$Correct == "Yes", "white", "white"),
              background = ifelse(summ$Correct == "Yes", "#27ae60",
                           ifelse(summ$Correct == "Partially", "#e67e22", "#e74c3c"))) %>%
  column_spec(3, width = "55%")
```

---

# Recommendations {#recommendations}

| Topic | Recommendation |
|-------|----------------|
| **Reference** | Use T2T-CHM13v2.0 for TE quantification — ~15% more locus-level detection |
| **Aligner** | Use STARsolo with `--outFilterMultimapNmax 100` (never CellRanger for TE work) |
| **Annotation** | Match the reference: use T2T RepeatMasker BED with T2T reference |
| **Thresholds** | Report ≥10 cells OR ≥50 counts as "robustly expressed" (~30% of annotated loci) |
| **Terminology** | Distinguish "annotated loci" (total matrix features) from "detected loci" (threshold-passing) |
| **Polymorphic TEs** | Young AluY* elements in T2T are individual-specific; some may be CHM13-only |
| **Comparison studies** | Always match aligner and multimapper settings before attributing differences to reference |

---

# Data Sources {#data-sources}

| Dataset | Repository | Key files |
|---------|------------|-----------|
| Patient snRNAseq | `jcbclffrd/soloTEmvp` master branch, commit `bbe49fd` | `comparison_T2T_vs_hg38_fair/` (April 21 2026) |
| Assembly liftover | `jcbclffrd/soloTEmvpWGsim` branch `analysis/hg38-t2t-liftover` | `analysis/liftover/results/` TSVs |
| hg38 RepeatMasker | UCSC `rmsk.txt.gz`, Oct 2022 | 5,683,690 total annotations |
| T2T RepeatMasker | `T2T-CHM13v2.0_RepeatMasker.bed`, project `references/` | 5,590,282 total annotations |
| Chain files | UCSC `hg38ToHs1.over.chain.gz`, `hs1ToHg38.over.chain.gz` | CrossMap v0.7.0 |
